---
title: "STAT 715 - Time Series Project"
author: "Daniel L."
date: "4/27/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, error = FALSE, message=FALSE, echo=FALSE}
### adf.test()
#rmarkdown::render("STAT_715_Presentation.Rmd")
library(tidyverse)
library(lubridate)
library(fable)
library(TSA)
#load forescasting package
library(fpp2)
library(ggThemeAssist)
library(kableExtra)
```



# Data Description

* This data was collected and updated by the NYC Mayor's Office of Operations and can be found on NYC Open Data
* This data focuses around 3 essential responding agencies
  + Emergency Medical Services
  + New York City Fire Department
  + New York City Police Department
* This data tracks the emergency incidents of the responding agencies as requested through the 911 phone system.
* The data is collected monthly, but reported weekly. The dates found in the data represent the first Monday of the week. So we will look at data for the following 7 days, beginning on the date reported.
* Our variable of interest is "Call to Agency Arrival." 
  + This attribute provides the average amount of time, in seconds, from when a 911 call is initiated, to when an agency arrives on scene.
* We will provide time series models for incidents that occurred from the 2015 to the end of 2019.


# Emergency Medical Services (EMS)

EMS workers respond to two different calls, "Life Threatening Medical Emergencies" and "Non-life Threatening Medical Emergencies"

```{r, echo = FALSE}
(EMS_table_final) %>% kbl(caption = "Number of Incidents Occured", align = 'c') %>% kable_classic(full_width = F) %>% kable_styling(bootstrap_options = "striped",full_width = FALSE, position = "center")
```

Let us compare the time series of the two types of incidents

####  Life-Threatening

```{r, echo = FALSE}
autoplot(ts_EMS_final_1) +
  ggtitle("Life Threatening Medical Emergencies: First Call to EMS arrival") +
  ylab("Seconds") + theme(panel.background = element_rect(fill = NA, 
    linetype = "dotted"), plot.background = element_rect(fill = "white", 
    colour = NA)) + theme(panel.grid.major = element_line(linetype = "dotted"), 
    panel.grid.minor = element_line(linetype = "dotted"), 
    panel.background = element_rect(fill = "antiquewhite"), 
    plot.background = element_rect(fill = "whitesmoke"))
```

####  Non-Life Threatening

```{r, fig.width=10, fig.height=7, echo = FALSE}
autoplot(ts_EMS_final_2) +
  ggtitle("Non-Life Threatening Medical Emergencies: First Call to EMS arrival") +
  ylab("Seconds") + theme(panel.grid.major = element_line(linetype = "dotted"), 
    panel.grid.minor = element_line(linetype = "dotted"), 
    panel.background = element_rect(fill = "antiquewhite"), 
    plot.background = element_rect(fill = "whitesmoke"))
```

Let us focus on Life Threatening Medical Emergencies.

### Let us difference this data once

The differenced time series uses the value of week 2 minus week 1 of 2015, week 3 minus week 2, so on. 

```{r, fig.width=10, fig.height=7, echo = FALSE}
autoplot(diff(ts_EMS_final_1)) +
  ggtitle("Differenced Threatening Medical Emergencies: First Call to EMS arrival") +
  ylab("Seconds") + theme(panel.grid.major = element_line(linetype = "dotted"), 
    panel.grid.minor = element_line(linetype = "dotted"), 
    panel.background = element_rect(fill = "antiquewhite"), 
    plot.background = element_rect(fill = "whitesmoke"))
```

The trend is removed, fluctuations do exist. Now, assuming this new series is stationary, let us investigate seasonality.

```{r, fig.width=10, fig.height=7, error = FALSE, echo = FALSE}
ggseasonplot(ts_EMS_final_1_diff) + ggtitle("Seasonal Plot: Change in First Call to EMS arrival") + ylab("Seconds") + theme(panel.grid.major = element_line(linetype = "dotted"), 
    panel.grid.minor = element_line(linetype = "dotted"), 
    panel.background = element_rect(fill = "antiquewhite"), 
    plot.background = element_rect(fill = "whitesmoke")) + labs(colour = "Year") + theme(legend.background = element_rect(fill = "whitesmoke"))
```

### These lag plots show our differenced time series $y_t$, plotted against $y_{t-k}$ for k $\in$ [1, 2, 3, 4]

```{r, fig.width=10, fig.height=7, echo = FALSE}
gglagplot(ts_EMS_final_1_diff, lags = 4, set.lags = 1:4, sesaonal = FALSE, do.lines = FALSE) + ggtitle("Lag Plot") + theme(legend.title = element_blank())
```

```{r, include="false", echo=FALSE}
gglagplot(ts_EMS_final_1, lags = 4, set.lags = 1:4, sesaonal = FALSE, do.lines = FALSE) + ggtitle("Lag Plot") + theme(legend.title = element_blank())
```

Here, we can not specify a positive or negative linear trend. 

```{r,echo=FALSE}
#ggAcf(ts_EMS_final_1_diff) + ggtitle("Autocorrelation Plot: Differenced Life Threatening Emergencies Time Series")
#Lags 1 and 2 go beyond the 95% confidence intervals.
```

### Let us model the differenced data with ARIMA

```{r}
print(summary(ts_EMS_final_1_fit_arima))
```

```{r}
checkresiduals(ts_EMS_final_1_fit_arima)
```
Autocorrelation measures the linear relationship between lagged values of a time series. This reinforces my claim that we do not see a strong negative or positive trend in the lagged time series

The number of lags used is 52, so our degrees of freedom for the residuals is 52-3-1 = 48. The upper tail $\chi^2$ evaluated at $\alpha = .05$ with 48 degrees of freedom is 65.171. Our test statistic, Q* = 41.052 and is $\leq$ 65.171. 

$H_0$: This ARIMA(3,1,1) model does not exhibit lack of fit vs $H_a$: This model exhibits lack of fit.

Out test statistic tells us that we fail to reject the null hypothesis, and conclude $H_0$. This model does not exhibit significant lack of fit.

## Begin forecasting

```{r, results = 'hide'}
forecast(ts_EMS_final_1_fit_arima, h = 52)
```

```{r, echo = FALSE, fig.width=10, fig.height=7}
autoplot(forecast(ts_EMS_final_1_fit_arima, h = 12)) + labs(title = "ARIMA(3,1,1) Forecast of Life Threatening Emergencies Data", y = "Seconds")
```

# NYPD

```{r, fig.width=10, fig.height=7, echo = FALSE}
NYPD_table_final %>% kbl(caption = "Number of Incidents Occured", align = 'c') %>% kable_classic(full_width = F) %>% kable_styling(bootstrap_options = "striped")
```

```{r, fig.width=10,fig.height=7, echo = FALSE}
ggplot(NYPD_data, aes(x = (Week_Start_Date), y = Call_to_Agency_Arrival)) + geom_line(aes(color = Final_Incident_Type), size = 1) + theme(panel.background = element_rect(fill = "antiquewhite1"), 
    plot.background = element_rect(fill = "whitesmoke"), 
    legend.key = element_rect(fill = "whitesmoke"), 
    legend.background = element_rect(fill = "whitesmoke")) +labs(title = "Time Series Plot:  Call to NYPD Arrival", 
    y = "Seconds")+labs(colour = "Incident Type", x = "Year") 
```

#### Let us plot the Critical Incidents time series

```{r, echo = FALSE,fig.width=10, fig.height=7}
autoplot(ts_NYPD_data_1) +
  ggtitle("Critical Incidents: First Call to NYPD arrival" )+
  ylab("Seconds") + theme(panel.grid.major = element_line(linetype = "dotted"), 
    panel.grid.minor = element_line(linetype = "dotted"), 
    panel.background = element_rect(fill = "antiquewhite"), 
    plot.background = element_rect(fill = "whitesmoke"))
```

# Let us plot differenced critical incidents time series

```{r, fig.width=10, fig.height=7, echo = FALSE}
ggseasonplot(diff(ts_NYPD_data_1)) + theme(panel.grid.major = element_line(linetype = "dotted"), 
    panel.grid.minor = element_line(linetype = "dotted"), 
    panel.background = element_rect(fill = "antiquewhite1"), 
    plot.background = element_rect(fill = "whitesmoke"), 
    legend.background = element_rect(fill = NA)) +labs(title = "Seasonal Plot of the Differenced NYPD Critical Incidents Data", y = "Seconds", colour = "Year")  + theme(plot.title = element_text(face = "plain"))
```

We will fit an ARIMA() model using the auto.arima() function and this is our output:

```{r}
print(summary(ts_NYPD_data_1_fit_arima))
```

```{r}
checkresiduals(ts_NYPD_data_1_fit_arima)
```

```{r}
qchisq(.95, 48)
```

The number of lags used is 52, so our degrees of freedom for the residuals is 52-4 = 48. The upper tail $\chi^2$ evaluated at $\alpha = .05$ with 48 degrees of freedom is 65.171. Our test statistic, Q* = 47.752 and is $\leq$ 65.171.

$H_0$: This ARIMA(0,1,3) with drift time series does not exhibit lack of fit. Vs $H_a$: This time series model exhibits lack of fit. 

We conclude $H_0$.

```{r, results='hide'}
forecast(ts_NYPD_data_1_fit_arima, h = 52);
```

```{r,echo = FALSE, fig.width=10, fig.height=7}
autoplot(forecast(ts_NYPD_data_1_fit_arima, h = 12)) + labs(title = "Forecasts from Critical Incidents ARIMA(0,1,3) with a Constant", y = "Seconds")
```

# FDNY

```{r,echo=FALSE, fig.width=10, fig.height=7}
FDNY_table_final2 %>% kable(caption = "Number of Incidents Occured", align = 'c') %>% kable_classic(full_width = F) %>% kable_styling(bootstrap_options = "striped")
```

```{r,fig.width=10,fig.height=7, echo=FALSE}
ggplot(FDNY_data, aes(x = Week_Start_Date, y = Call_to_Agency_Arrival)) + 
  geom_line(aes(color = Final_Incident_Type), size = 1) +
  scale_color_manual(values = c("#00AFBB", "#E7B800", "green", "black")) +
  theme_minimal()+labs(title = "Time Series Plot: Call to FDNY Arrival", 
    y = "Seconds") + theme(panel.background = element_rect(fill = "antiquewhite"), 
    plot.background = element_rect(fill = "whitesmoke"), 
    legend.background = element_rect(fill = "gray90")) + theme(panel.grid.major = element_line(linetype = "twodash"), 
    panel.grid.minor = element_line(linetype = "twodash")) + theme(panel.grid.major = element_line(linetype = "dotted"), panel.grid.minor = element_line(linetype = "dotted")) + theme(legend.background = element_rect(fill = "whitesmoke")) + labs(x = "Year", colour = "Incident Type")
```

There appears to be a strange spike in response times for three of the incident types. The data dictionary provided advises that if the volume of incidents surpasses a point, response times will degenerate. 

#### Let us isolate the structural fire time series 

```{r, fig.width=10, fig.height=7, echo = FALSE}
autoplot(ts_FDNY_data_1) +
  ggtitle("Structural Fires: First Call to FDNY arrival" ) +
  ylab("Seconds") + theme(panel.grid.major = element_line(linetype = "dotted"), 
    panel.grid.minor = element_line(linetype = "dotted"), 
    panel.background = element_rect(fill = "antiquewhite"), 
    plot.background = element_rect(fill = "whitesmoke"))
```


#### Let us plot the differenced Structural Fires time series:

```{r, echo = FALSE, fig.width=10,fig.height=7}
ggseasonplot(diff(ts_FDNY_data_1)) +labs(title = "Time Series Difference Plot: First Call to FDNY Arrival of Structural Fires", 
    y = "Seconds") + theme(panel.background = element_rect(fill = "antiquewhite1"), 
    plot.background = element_rect(fill = "whitesmoke"), 
    legend.background = element_rect(fill = "gray90")) + theme(panel.grid.major = element_line(linetype = "twodash"), 
    panel.grid.minor = element_line(linetype = "twodash")) + theme(panel.grid.major = element_line(linetype = "dotted"), 
    panel.grid.minor = element_line(linetype = "dotted")) + theme(legend.background = element_rect(fill = "whitesmoke")) + labs(colour = "Year")
```

### Fitting an ARIMA Model to the differenced Structural Fires data

```{r}
print(summary(ts_FDNY_data_fit))
```
```{r, fig.width=10, fig.height=7}
checkresiduals(ts_FDNY_data_fit)
```

```{r}
qchisq(0.95,50)
```

Our Q-statistic is again, less than $\chi^2$ at $\alpha = 0.05$ with 50 degrees of freedom. We fail to reject the null hypothesis and conclude $H_0$

### Begin forecasting

```{r, results='hide'}
forecast(ts_FDNY_data_fit)
```

```{r,echo = FALSE, fig.width=10, fig.height=7}
autoplot(forecast(diff(ts_FDNY_data_1))) + theme(panel.grid.major = element_line(linetype = "dashed"), 
    panel.grid.minor = element_line(linetype = "dashed"), 
    panel.background = element_rect(fill = "gray96")) +labs(title = "MA(1) with Drift Forecast of Differenced Structural Fires Data", 
    y = "Seconds")
```

```{r,echo=FALSE,fig.width=10, fig.height=7}
autoplot(forecast(ts_FDNY_data_fit), h = 52) + labs(title = "MA(1) with a Constant Forecast of Structural Fires Data", y = "Seconds")
```

# Any questions? I would appreciate any constructive criticism. Thank you!
