---
title: "STAT 715 Presentation"
author: "Daniel L."
date: "4/19/2021"
output: html_document
---

rmarkdown::render("STAT_715_Presentation.Rmd")

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, error = FALSE, message=FALSE, echo=FALSE}
library(tidyverse)
library(lubridate)
library(fable)
library(TSA)
#load forescasting package
library(fpp2)
library(ggThemeAssist)
library(kableExtra)
```

# Emergency Medical Services (EMS)

Begin by introducing its services and the data that comes with it

EMS workers respond to two different calls, "Life Threatening Medical Emergencies" and "Non-life Threatening Medical Emergencies"

-Create a table of counts

-Compare the time series

```{r, echo = FALSE}
EMS_table_final %>% kbl(caption = "Number of Incidents Occured") %>% kable_classic(full_width = F) %>% kable_styling(position = "center")
```

Let us compare the time series of the two types of incidents

### Life-Threatening

```{r, echo = FALSE}
autoplot(ts_EMS_final_1) +
  ggtitle("Life Threatening Medical Emergencies: First Call to EMS arrival") +
  ylab("Seconds") + theme(panel.background = element_rect(fill = NA, 
    linetype = "dotted"), plot.background = element_rect(fill = "white", 
    colour = NA))
```

### Non-Life Threatening

```{r, fig.width=10, fig.height=7, echo = FALSE}
autoplot(ts_EMS_final_2) +
  ggtitle("Non-Life Threatening Medical Emergencies: First Call to EMS arrival") +
  ylab("Seconds") + theme(panel.background = element_rect(fill = NA, 
    linetype = "dotted"), plot.background = element_rect(fill = "white", 
    colour = NA))
```

Let us focus on Life Threatening Medical Emergencies.

### Let us difference this data once

The differenced time series uses the value of week 2 minus week 1 of 2015, week 3 minus week 2, so on. 

```{r, fig.width=10, fig.height=7, echo = FALSE}
ts_EMS_final_1_diff <- diff(ts_EMS_final_1)
autoplot(ts_EMS_final_1_diff) +
  ggtitle("Differenced Life Threatening Medical Emergencies: First Call to EMS arrival" )+
  ylab("Seconds")
```

The trend is removed, fluctuations do exist. Now, assuming this new series is stationary, let us investigate seasonality.

```{r, fig.width=10, fig.height=7, error = FALSE, echo = FALSE}
ggseasonplot(ts_EMS_final_1_diff) + ggtitle("Seasonal Plot: Change in First Call to EMS arrival") + ylab("Seconds")+ theme_minimal()
```


```{r, fig.width=10, fig.height=7, echo = FALSE}
gglagplot(ts_EMS_final_1_diff, lags = 4, set.lags = 1:4, sesaonal = FALSE, do.lines = FALSE) + ggtitle("Lag Plot") + theme(legend.title = element_blank())
```

Here, we can not specify a positive or negative linear trend. 

```{r,echo=FALSE}
#ggAcf(ts_EMS_final_1_diff) + ggtitle("Autocorrelation Plot: Differenced Life Threatening Emergencies Time Series")
```

Lags 1 and 2 go beyond the 95% confidence intervals.

### Let us model the differenced data with ARIMA

```{r}
print(summary(ts_EMS_final_1_fit_arima))
```

```{r}
checkresiduals(ts_EMS_final_1_fit_arima)
```

The number of lags used is 52, so our degrees of freedom for the residuals is 52-3-1 = 48. The upper tail $\chi^2$ evaluated at $\alpha = .05$ with 48 degrees of freedom is 65.171. Our test statistic, Q* = 41.052 and is $\leq$ 65.171. 

$H_0$: This ARIMA(3,1,1) model does not exhibit lack of fit vs $H_a$: This model exhibits lack of fit.

Out test statistic tells us that we fail to reject the null hypothesis, and conclude $H_0$. This model does not exhibit has significant lack of fit.

## Begin forecasting

```{r, results = 'hide'}
forecast(ts_EMS_final_1_fit_arima, h = 52)
```

```{r, fig.width=10, fig.height=7}
autoplot(forecast(ts_EMS_final_1_fit_arima, h = 52))
```

# NYPD

```{r, fig.width=10, fig.height=7, echo = FALSE}
NYPD_table_final %>% kbl(caption = "Number of Incidents Occured") %>% kable_classic(full_width = F) %>% kable_styling(position = "center")
```

```{r, fig.width=10,fig.height=7, echo = FALSE}
ggplot(NYPD_data, aes(x = (Week_Start_Date), y = Call_to_Agency_Arrival)) + geom_line(aes(color = Final_Incident_Type), size = 1) + theme(panel.background = element_rect(fill = "antiquewhite1"), 
    plot.background = element_rect(fill = "whitesmoke"), 
    legend.key = element_rect(fill = "whitesmoke"), 
    legend.background = element_rect(fill = "whitesmoke")) +labs(title = "Time Series Plot:  Call to NYPD Arrival", 
    y = "Seconds")+labs(colour = "Incident Type", x = "Year") 
```


We will fit an ARIMA() model using the 

```{r, fig.width=10, fig.height=7, echo = FALSE}
ggseasonplot(diff(ts_NYPD_data_1)) + theme(panel.grid.major = element_line(linetype = "dotted"), 
    panel.grid.minor = element_line(linetype = "dotted"), 
    plot.title = element_text(face = "bold"), 
    panel.background = element_rect(fill = "antiquewhite1"), 
    plot.background = element_rect(fill = "whitesmoke"), 
    legend.background = element_rect(fill = NA)) +labs(title = "Seasonal Plot of the Differenced NYPD Critical Incidents Data", 
    y = "Seconds", colour = "Year") 
```

```{r}
print(summary(ts_NYPD_data_1_fit_arima))
```

```{r}
checkresiduals(ts_NYPD_data_1_fit_arima)
```

```{r}
qchisq(.95, 48)
```

```{r, results='hide'}
forecast(ts_NYPD_data_1_fit_arima, h = 52);
```
```{r, fig.width=10, fig.height=7}
autoplot(forecast(ts_NYPD_data_1_fit_arima, h = 52)) + labs(y = "Seconds")
```

# FDNY

```{r, echo=FALSE}
FDNY_table_final %>% kable(caption = "Number of Incidents Occured") %>% kable_classic(full_width = F) %>% kable_styling(position = "center")
```

```{r,fig.width=10,fig.height=7, echo=FALSE}
ggplot(FDNY_data, aes(x = Week_Start_Date, y = Call_to_Agency_Arrival)) + 
  geom_line(aes(color = Final_Incident_Type), size = 1) +
  scale_color_manual(values = c("#00AFBB", "#E7B800", "green", "black")) +
  theme_minimal()+labs(title = "Time Series Plot: Call to FDNY Arrival", 
    y = "Seconds") + theme(panel.background = element_rect(fill = "antiquewhite1"), 
    plot.background = element_rect(fill = "whitesmoke"), 
    legend.background = element_rect(fill = "gray90")) + theme(panel.grid.major = element_line(linetype = "twodash"), 
    panel.grid.minor = element_line(linetype = "twodash")) + theme(panel.grid.major = element_line(linetype = "dotted"), 
    panel.grid.minor = element_line(linetype = "dotted")) + theme(legend.background = element_rect(fill = "whitesmoke")) + labs(x = "Year", colour = "Incident Type")
```

```{r, fig.width=10, fig.height=7, echo = FALSE}
autoplot(ts_FDNY_data_1) +
  ggtitle("Structural Fires: First Call to FDNY arrival" )+
  ylab("Seconds") + theme(panel.grid.major = element_line(linetype = "dotted"), 
    panel.grid.minor = element_line(linetype = "dotted"), 
    panel.background = element_rect(fill = "antiquewhite"), 
    plot.background = element_rect(fill = "whitesmoke"))

```

